---
title: "ADA Accessibility Prediction"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r default_setting, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
library(tidyverse)
library(sf)
library(purrr)
library(tidyverse)
library(dplyr)
library(janitor)
library(here)
library(reader)
library(nnet)
library(knitr)
```

```{r retreiving_cleaned_merged_data, echo = FALSE, message = FALSE, warning = FALSE}
restroom_subway_join = read_csv(here::here("./Data/cleaned_subway_restroom_data.csv"))
```


```{r predicting_ada_prep, echo = FALSE, message = FALSE, warning = FALSE}
# Cleaning merged data
restroom_subway_data = restroom_subway_join %>%
  dplyr::select(
    restroom_accessibility,
      # Dependent variable (ADA accessibility)
    restroom_latitude, restroom_longitude,
      # Restroom location features
    restroom_changing_stations, restroom_open,
      # Restroom other features
    station_latitude, station_longitude,
      # Subway station location features
    entrance_type, staffing
      # Subway station other features
  ) %>%
  drop_na() %>% 
  mutate(
    restroom_accessibility = as.factor(as.character(restroom_accessibility)), 
      # str(restroom_subway_data$restroom_accessibility)
        # confirming if restroom_accessibility is no longer ordered
    restroom_accessibility = relevel(restroom_accessibility, ref = "Not Accessible"),
    restroom_open = as.factor(restroom_open),
    entrance_type = as.factor(entrance_type),
    staffing = as.factor(staffing)
  )
```

```{r predicting_model, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
# Multinomial logistic regression model
ada_model = multinom(restroom_accessibility ~ restroom_latitude + restroom_longitude +
                       restroom_changing_stations + restroom_open +
                       station_latitude + station_longitude + 
                       entrance_type + staffing,
                     data = restroom_subway_data)

# Check the model summary
summary(ada_model)
```

**Model Summary Table**
```{r model_table, echo = FALSE, message = FALSE, warning = FALSE}
table = head(broom::tidy(ada_model), 31)
knitr::kable(table)
```
Based on this output, we see that the restroom location, ramp entrance types and partial staffs at train stations are highly significant predictors of a public restroom being full ADA accessible. For public restrooms that are partially ADA accessible, restroom location, ramp entrance types and no staffs members at train stations are strong predictors. Variables such as easement entrance type, elevator entrance type, and stairs entrance type for full ADA accesiblity have high p-values, indicating weak evidence of their association with a restroom being full ADA accesible. Potential multicollinearity or insufficient data may have resulted in very large or null statistical value on escalaltor entrance type and ramp entrance type.

```{r testing_model, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
predicted_accessibility = predict(ada_model, restroom_subway_data)

# Confusion Matrix to evaluate the model's performance
table(Predicted = predicted_accessibility, Actual = restroom_subway_data$restroom_accessibility)

# Calculate accuracy
accuracy = mean(predicted_accessibility == restroom_subway_data$restroom_accessibility) * 100
accuracy = print(paste(accuracy, "%"))
```
**Accuracy:**

When testing the prediction model on the public restroom and train station data, there is an approximate of `r accuracy` accuracy, which may be due to outliers, missing data, or unequal distribution of data among different catergories of a variable.