---
title: "Hot Spot Analysis (Getis-Ord Gi)"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r default_setting, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
library(sf)         
library(dplyr)       
library(leaflet)     
library(knitr)       
library(readr)      
library(janitor)    
library(spdep)      
library(here) 
library(ggplot2)
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r clean_restroom_data, echo = FALSE, message = FALSE, warning = FALSE}
# Import data
restroom_cleaned = read_csv(here::here("./Data/Public_Restrooms_20241203.csv")) %>% 
  janitor::clean_names()
subway_df = read_csv(here::here("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")) %>% 
  janitor::clean_names()

# Clean restroom data
restroom_cleaned = restroom_cleaned %>% 
  select(
    -website, -operator, -hours_of_operation 
  ) %>% 
  rename(
    restroom_latitude = latitude,
    restroom_longitude = longitude,
    restroom_location = location
  ) %>% 
  mutate(
    restroom_latitude = as.numeric(restroom_latitude),
    restroom_longitude = as.numeric(restroom_longitude),
    restroom_location = st_as_sfc(restroom_location), #convert to point
    restroom_open = factor(
      open,
      levels = c("Future", "Seasonal", "Year Round"),
      ordered = TRUE
    ),
    restroom_accessibility = factor(
      accessibility,
      levels = c("Not Accessible", "Partially Accessible", "Fully Accessible"),
      ordered = TRUE
    ),
    restroom_changing_stations = case_when(
      changing_stations %in% c("Yes, in single-stall all gender restroom only",
                                "Yes, in women's restroom only",
                                "Yes") ~ 1,
      changing_stations == "No" ~ 0
    ),
    restroom_status = case_when(
      status %in% c("Not Operational",
                    "Closed for Construction",
                    "Closed") ~ 0,
      status == "Operational" ~ 1
    )
  ) 

# Convert dataframe to sf for spatial operations
restroom_sf = st_sf(restroom_cleaned, crs = 4326)
restroom_df = restroom_cleaned %>% 
  filter(location_type == 'Transit')
```


```{r clean_subway_data, echo = FALSE, message = FALSE, warning = FALSE}
# Load the dataset
subway_df = read_csv(here::here("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv"))

# Clean the dataset
subway_cleaned = subway_df %>% 
  janitor::clean_names() %>% 
  select( -ada_notes, -staff_hours, -north_south_street, -east_west_street, -corner) %>%
  rename(
    subway_latitude = entrance_latitude,
    subway_longitude = entrance_longitude,
    subway_location = entrance_location
  ) %>% 
  drop_na(subway_latitude, subway_longitude) %>%  # Drop rows with missing coordinates to avoid errors
  st_as_sf(coords = c("subway_longitude", "subway_latitude"), crs = 4326) %>% 
  mutate(
    entrance_type = case_when(
      entrance_type == "Elevator" ~ 1,
      entrance_type == "Ramp" ~ 2,
      entrance_type == "Escalator" ~ 3,
      entrance_type == "Walkway" ~ 4,
      entrance_type == "Door" ~ 5,
      entrance_type == "Easement" ~ 6,
      entrance_type == "Stair" ~ 7,
      TRUE ~ NA_real_  # Handle unexpected values as NA
    )
  ) %>% 
  mutate(
    staffing = factor(
      staffing,
      levels = c("NONE", "Spc Ev", "PART", "FULL"),
      ordered = TRUE
    )
  )

# Convert subway data to sf object
subway_sf = st_sf(subway_cleaned, crs = 4326)

subway_with_restrooms = st_join(subway_sf, restroom_sf, join = st_nearest_feature)
```

Hot Spot Analysis using the Getis-Ord Gi* statistic is a spatial statistical method used to identify statistically significant clusters (hot spots or cold spots) of high or low values in geographic data. This technique helps determine areas where a phenomenon (e.g., restroom accessibility) is unusually concentrated or sparse.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
# Step 1: Load ZIP Code Shapefile
zip_shapes <- st_read("./NYC/zipcode.shp") %>%
  st_transform(crs = 4326) %>%
  rename(zip_code = modzcta)
```


```{r Getis-Ord_Gi, echo = FALSE, message = FALSE, warning = FALSE}

# Step 2: Convert Restrooms to sf Object
restroom_sf <- st_as_sf(
  restroom_cleaned,
  coords = c("restroom_longitude", "restroom_latitude"),  # Replace with your column names
  crs = 4326
)

# Step 3: Spatial Join to Assign ZIP Codes
restroom_sf <- st_transform(restroom_sf, crs = st_crs(zip_shapes))
restroom_with_zip <- st_join(restroom_sf, zip_shapes, join = st_within)

# Step 4: Count Restrooms in Each ZIP Code
restroom_density <- restroom_with_zip %>%
  st_drop_geometry() %>%  # Drop geometry for simplification
  group_by(zip_code) %>%
  summarize(restroom_count = n(), .groups = "drop")

# Step 5: Join Restroom Counts with ZIP Code Shapefile
zip_shapes <- zip_shapes %>%
  mutate(area_km2 = st_area(geometry) / 1e6) %>%  # Area in square kilometers
  left_join(restroom_density, by = "zip_code") %>%
  mutate(
    restroom_count = replace_na(restroom_count, 0),  # Fill missing values with 0
    density = restroom_count / as.numeric(area_km2)  # Density (restrooms per km²)
  )

# Step 6: Define Neighbors and Spatial Weights
zip_nb <- poly2nb(zip_shapes, snap = 0.001)  # Neighbors list with snapping
zip_weights <- nb2listw(zip_nb, style = "W", zero.policy = TRUE)  # Row-standardized weights

# Step 7: Perform Getis-Ord Gi* Analysis
gi_star <- localG(zip_shapes$density, zip_weights)
zip_shapes <- zip_shapes %>%
  mutate(gi_star = as.numeric(gi_star))  # Convert to numeric for mapping

# Step 8: Define a Custom Legend Function
addLegendCustom <- function(map, colors, labels, title, position = "bottomright") {
  colorAdditions <- paste0(
    '<i style="background:', colors, ';width:15px;height:15px;display:inline-block;margin-right:5px"></i>'
  )
  labelAdditions <- paste0('<span style="margin-left:10px">', labels, '</span>')
  combined <- paste0(colorAdditions, labelAdditions, collapse = "<br>")
  legendHTML <- paste0(
    '<div style="margin-left: 15px;"><strong>', title, '</strong><br>',
    combined,
    '</div>'
  )
  addControl(map, html = legendHTML, position = position)
}

# Step 9: Create the Interactive Map

# Custom Legend Function to Include NA on a Separate Line
addLegendCustom <- function(map, pal, values, title, na_label, na_color, position = "bottomright") {
  # Create the gradient legend
  colorAdditions <- paste0(
    '<i style="background:', pal(seq(min(values, na.rm = TRUE), max(values, na.rm = TRUE), length.out = 5)), 
    ';width:15px;height:15px;display:inline-block;margin-right:5px"></i>',
    c(round(seq(min(values, na.rm = TRUE), max(values, na.rm = TRUE), length.out = 5), 2))
  )
  
  # Add NA entry at the bottom
  naEntry <- paste0(
    '<div style="margin-top: 10px;">',
    '<i style="background:', na_color, ';width:15px;height:15px;display:inline-block;margin-right:5px"></i>',
    na_label,
    '</div>'
  )
  
  # Combine all legend entries
  legendHTML <- paste0(
    '<div style="margin-left: 15px;"><strong>', title, '</strong><br>',
    paste(colorAdditions, collapse = "<br>"), 
    naEntry,
    '</div>'
  )
  
  # Add the custom legend to the map
  addControl(map, html = legendHTML, position = position)
}

# Define a color palette for Gi* Z-scores
pal <- colorNumeric(
  palette = "Reds",
  domain = zip_shapes$gi_star,
  na.color = "gray"  # Color for NA values
)

# Create the interactive map
leaflet(data = zip_shapes) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal(gi_star),
    weight = 0.5,
    color = "black",
    fillOpacity = 0.7,
    popup = ~paste0(
      "<b>ZIP Code:</b> ", zip_code, "<br>",
      "<b>Restrooms:</b> ", restroom_count, "<br>",
      "<b>Density:</b> ", round(density, 2), " restrooms/km²<br>",
      "<b>Gi* Z-Score:</b> ", round(gi_star, 2)
    )
  ) %>%
  addLegendCustom(
    pal = pal,
    values = zip_shapes$gi_star,
    title = "Getis-Ord Gi*<br>Hot Spot Analysis",
    na_label = "No Data (NA)",
    na_color = "gray",
    position = "bottomright"
  )

```
The map identifies significant hot spots (red areas) for restroom density, particularly concentrated in Manhattan, indicating clusters of high restroom accessibility in these ZIP codes.

Neutral areas (light colors) suggest no significant clustering, reflecting average restroom accessibility levels across parts of Queens, Brooklyn, and the Bronx.

Cold spots or areas with no data (gray) are observed in regions like Staten Island and some peripheral areas, highlighting gaps in restroom density or missing data requiring further investigation.


```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Calculate distances between restrooms and subway entrances
subway_restroom_dist <- st_distance(subway_sf, restroom_sf)

# Identify underserved areas (e.g., areas without restrooms within 500m)
underserved_subway <- subway_sf %>%
  mutate(min_distance = apply(subway_restroom_dist, 1, min)) %>%
  filter(min_distance > 500)
# Select a sample of 5 underserved subway stations
underserved_sample <- underserved_subway %>%
  st_drop_geometry() %>%  # Drop geometry for table creation
  select(station_name, line, station_latitude, station_longitude, min_distance) %>%
  slice_head(n = 5)  # Get the first 5 rows

# Create a table
library(knitr)
underserved_sample %>%
  mutate(
    min_distance = round(min_distance, 2)  # Round distances for readability
  ) %>%
  kable(
    caption = "Sample of 5 Underserved Subway Stations",
    col.names = c("Station Name", "Line", "Latitude", "Longitude", "Min Distance to Restroom (m)"),
    digits = 2
  )
```
There are 272 stations without restrooms within 500m. 


```{r general, echo=FALSE, message=FALSE, warning=FALSE}
# Define a custom legend function
addLegendCustom <- function(map, pal, values, title, na_label, na_color, position = "bottomright") {
  # Create the gradient legend
  colorAdditions <- paste0(
    '<i style="background:', pal(seq(min(values, na.rm = TRUE), max(values, na.rm = TRUE), length.out = 5)), 
    ';width:15px;height:15px;display:inline-block;margin-right:5px"></i>',
    c(round(seq(min(values, na.rm = TRUE), max(values, na.rm = TRUE), length.out = 5), 2))
  )
  
  # Add NA entry at the bottom
  naEntry <- paste0(
    '<div style="margin-top: 10px;">',
    '<i style="background:', na_color, ';width:15px;height:15px;display:inline-block;margin-right:5px"></i>',
    na_label,
    '</div>'
  )
  
  # Combine all legend entries
  legendHTML <- paste0(
    '<div style="margin-left: 15px;"><strong>', title, '</strong><br>',
    paste(colorAdditions, collapse = "<br>"), 
    naEntry,
    '</div>'
  )
  
  # Add the custom legend to the map
  addControl(map, html = legendHTML, position = position)
}

# Define the interactive map function
interactive_map <- function(restroom_cleaned, subway_cleaned, underserved_subway, zip_shapes) {
  
  # Define a custom yellow triangle icon for restrooms
  yellow_triangle_icon <- makeIcon(
    iconUrl = "https://raw.githubusercontent.com/pointhi/leaflet-color-markers/master/img/marker-icon-yellow.png",
    iconWidth = 10,
    iconHeight = 10
  )
  
  # Define a color palette for Getis-Ord Gi* Z-scores
  pal <- colorNumeric(
    palette = "Reds",
    domain = zip_shapes$gi_star,
    na.color = "gray"  # Color for NA values
  )
  
  leaflet() %>%
    # Add CartoDB Positron base layer
    addProviderTiles(providers$CartoDB.Positron) %>%
    
    # Add restroom locations as a layer with yellow triangle icons
    addMarkers(
      data = restroom_cleaned,
      lng = ~restroom_longitude,
      lat = ~restroom_latitude,
      group = "Restrooms",
      label = ~facility_name,
      popup = ~paste0(
        "<b>Restroom:</b> ", facility_name, "<br>",
        "<b>Location:</b> ", restroom_location
      ),
      icon = yellow_triangle_icon
    ) %>%
    
    # Add subway locations as a layer
    addCircleMarkers(
      data = subway_cleaned,
      lng = ~station_longitude,
      lat = ~station_latitude,
      group = "Subway Stations",
      label = ~station_name,
      popup = ~paste0(
        "<b>Station:</b> ", station_name, "<br>",
        "<b>Line:</b> ", line
      ),
      color = "blue",
      fillOpacity = 0.7,
      radius = 2.5
    ) %>%
    
    # Add underserved subway stations as a layer
    addCircleMarkers(
      data = underserved_subway,
      lng = ~station_longitude,
      lat = ~station_latitude,
      group = "Underserved Stations",
      label = ~station_name,
      popup = ~paste0(
        "<b>Station:</b> ", station_name, "<br>",
        "<b>Line:</b> ", line, "<br>",
        "<b>Distance to Restroom:</b> ", round(min_distance, 2), " meters"
      ),
      color = "red",
      fillOpacity = 0.7,
      radius = 2.5
    ) %>%
    
    # Add Getis-Ord Gi* Z-scores as a layer
    addPolygons(
      data = zip_shapes,
      fillColor = ~pal(gi_star),
      weight = 0.5,
      color = "black",
      fillOpacity = 2,
      group = "Getis-Ord Gi*",
      popup = ~paste0(
        "<b>ZIP Code:</b> ", zip_code, "<br>",
        "<b>Restrooms:</b> ", restroom_count, "<br>",
        "<b>Density:</b> ", round(density, 2), " restrooms/km²<br>",
        "<b>Gi* Z-Score:</b> ", round(gi_star, 2)
      )
    ) %>%
    
    # Add layer control
    addLayersControl(
      baseGroups = c("CartoDB"),
      overlayGroups = c("Restrooms", "Subway Stations", "Underserved Stations", "Getis-Ord Gi*"),
      options = layersControlOptions(collapsed = FALSE)
    ) %>%
    
    # Add legend for Getis-Ord Gi* Z-scores
    addLegendCustom(
      map = .,
      pal = pal,
      values = zip_shapes$gi_star,
      title = "Getis-Ord Gi*<br>Hot Spot Analysis",
      na_label = "No Data (NA)",
      na_color = "gray",
      position = "bottomright"
    )
}

# Render the interactive map
interactive_map(restroom_cleaned, subway_cleaned, underserved_subway, zip_shapes)



```
This interactive map displays the locations of restrooms and subway stations. You can explore the map by toggling layers for restrooms, subway stations, underserved stations, and Getis-Ord Gi* results, zooming in and out, and hovering over icons to view detailed information about the name and location of restrooms and stations.











