---
title: "Hotspot"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

### Data Cleaning

```{r default_setting, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
library(tidyverse)
library(sf)
library(ggplot2)
library(purrr)
library(tidyverse)
library(dplyr)
library(janitor)
library(here)
library(leaflet)
library(reader)
library(spatstat)

library(tidygeocoder)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r clean_restroom_data, echo = FALSE, message = FALSE, warning = FALSE}
# Import data
restroom_cleaned = read_csv(here::here("./Data/Public_Restrooms_20241203.csv")) %>% 
  janitor::clean_names()
subway_df = read_csv(here::here("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")) %>% 
  janitor::clean_names()

# Clean restroom data
restroom_cleaned = restroom_cleaned %>% 
  select(
    -website, -operator, -hours_of_operation 
  ) %>% 
  rename(
    restroom_latitude = latitude,
    restroom_longitude = longitude,
    restroom_location = location
  ) %>% 
  mutate(
    restroom_latitude = as.numeric(restroom_latitude),
    restroom_longitude = as.numeric(restroom_longitude),
    restroom_location = st_as_sfc(restroom_location), #convert to point
    restroom_open = factor(
      open,
      levels = c("Future", "Seasonal", "Year Round"),
      ordered = TRUE
    ),
    restroom_accessibility = factor(
      accessibility,
      levels = c("Not Accessible", "Partially Accessible", "Fully Accessible"),
      ordered = TRUE
    ),
    restroom_changing_stations = case_when(
      changing_stations %in% c("Yes, in single-stall all gender restroom only",
                                "Yes, in women's restroom only",
                                "Yes") ~ 1,
      changing_stations == "No" ~ 0
    ),
    restroom_status = case_when(
      status %in% c("Not Operational",
                    "Closed for Construction",
                    "Closed") ~ 0,
      status == "Operational" ~ 1
    )
  ) 

# Convert dataframe to sf for spatial operations
restroom_sf = st_sf(restroom_cleaned, crs = 4326)
restroom_df = restroom_cleaned %>% 
  filter(location_type == 'Transit')
```


```{r clean_subway_data, echo = FALSE, message = FALSE, warning = FALSE}
# Load the dataset
subway_df = read_csv(here::here("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv"))

# Clean the dataset
subway_cleaned = subway_df %>% 
  janitor::clean_names() %>% 
  select( -ada_notes, -staff_hours, -north_south_street, -east_west_street, -corner) %>%
  rename(
    subway_latitude = entrance_latitude,
    subway_longitude = entrance_longitude,
    subway_location = entrance_location
  ) %>% 
  drop_na(subway_latitude, subway_longitude) %>%  # Drop rows with missing coordinates to avoid errors
  st_as_sf(coords = c("subway_longitude", "subway_latitude"), crs = 4326) %>% 
  mutate(
    entrance_type = case_when(
      entrance_type == "Elevator" ~ 1,
      entrance_type == "Ramp" ~ 2,
      entrance_type == "Escalator" ~ 3,
      entrance_type == "Walkway" ~ 4,
      entrance_type == "Door" ~ 5,
      entrance_type == "Easement" ~ 6,
      entrance_type == "Stair" ~ 7,
      TRUE ~ NA_real_  # Handle unexpected values as NA
    )
  ) %>% 
  mutate(
    staffing = factor(
      staffing,
      levels = c("NONE", "Spc Ev", "PART", "FULL"),
      ordered = TRUE
    )
  )

# Convert subway data to sf object
subway_sf = st_sf(subway_cleaned, crs = 4326)

subway_with_restrooms = st_join(subway_sf, restroom_sf, join = st_nearest_feature)
```


### Hotspot Analysis

underserverd subway stations
```{r}
# Calculate distances between restrooms and subway entrances
subway_restroom_dist <- st_distance(subway_sf, restroom_sf)

# Identify underserved areas (e.g., areas without restrooms within 500m)
underserved_subway <- subway_sf %>%
  mutate(min_distance = apply(subway_restroom_dist, 1, min)) %>%
  filter(min_distance > 500)

```
There are 272 stations without restrooms within 500m. 

calculate the density of restroom in each zipcode
```{r}
# Load the shapefile
zip_shapes <- st_read("./NYC/zipcode.shp") %>%
  st_transform(zip_shapes, crs = 4326)

# Calculate area in square kilometers
zip_shapes <- zip_shapes %>%
  mutate(area_km2 = st_area(geometry) / 1e6)  # Convert from square meters to square kilometers

# Count the number of restrooms in each ZIP code
restroom_density <- restroom_cleaned %>%
  group_by(zip_code) %>%
  summarize(restroom_count = n(), .groups = "drop")

# Join the restroom counts with the shapefile
zip_shapes <- zip_shapes %>%
  left_join(restroom_density, by = c("GEOID" = "zip_code")) %>%
  mutate(
    restroom_count = replace_na(restroom_count, 0),  # Fill missing values with 0
    density = restroom_count / as.numeric(area_km2)  # Density (restrooms per km²)
  )

```


```{r}
# Function to create the interactive map
interactive_map <- function(restroom_cleaned, subway_cleaned, underserved_subway) {
  # Define a custom yellow triangle icon for restrooms
  yellow_triangle_icon <- makeIcon(
    iconUrl = "https://raw.githubusercontent.com/pointhi/leaflet-color-markers/master/img/marker-icon-yellow.png",
    iconWidth = 20,
    iconHeight = 20
  )
  
  leaflet() %>%
    # Add CartoDB Positron base layer
    addProviderTiles(providers$CartoDB.Positron) %>%
    
    # Add restroom locations as a layer with yellow triangle icons
    addMarkers(
      data = restroom_cleaned,
      lng = ~restroom_longitude,
      lat = ~restroom_latitude,
      group = "Restrooms",
      label = ~facility_name,
      popup = ~paste0(
        "<b>Restroom:</b> ", facility_name, "<br>",
        "<b>Location:</b> ", restroom_location
      ),
      icon = yellow_triangle_icon
    ) %>%
    
    # Add subway locations as a layer
    addCircleMarkers(
      data = subway_cleaned,
      lng = ~station_longitude,
      lat = ~station_latitude,
      group = "Subway Stations",
      label = ~station_name,
      popup = ~paste0(
        "<b>Station:</b> ", station_name, "<br>",
        "<b>Line:</b> ", line
      ),
      color = "blue",
      fillOpacity = 0.7,
      radius = 5
    ) %>%
    
    # Add subway locations as a layer
    addCircleMarkers(
      data = underserved_subway,
      lng = ~station_longitude,
      lat = ~station_latitude,
      group = "Underserved Stations",
      label = ~station_name,
      popup = ~paste0(
        "<b>Station:</b> ", station_name, "<br>",
        "<b>Line:</b> ", line
      ),
      color = "red",
      fillOpacity = 0.7,
      radius = 5
    ) %>%
    
    # Add layer control
    addLayersControl(
      baseGroups = c("CartoDB"),
      overlayGroups = c("Restrooms", "Subway Stations", "Underserved Stations"),
      options = layersControlOptions(collapsed = FALSE)
    )
}

# Render the map
interactive_map(restroom_cleaned, subway_cleaned, underserved_subway)
```
This interactive map contains the location of restrooms and subway stations. You can play around of this map by select and deselect the layer of restrooms and subway station layer, zoom in and zoom out the map, and moving your cursor to each icon to discover the name and location of restrooms and stations. 

## Kernel Density Estimation (KDE)
```{r}
# Convert the data to a point pattern object
# Ensure the data contains 'restroom_longitude' and 'restroom_latitude'
restroom_points <- as(restroom_sf, "Spatial")  # Convert sf object to Spatial object
window <- owin(xrange = range(restroom_points@coords[,1]), 
               yrange = range(restroom_points@coords[,2]))

restroom_ppp <- ppp(
  x = restroom_points@coords[,1], 
  y = restroom_points@coords[,2], 
  window = window
)

# Kernel density estimation
kden <- density(restroom_ppp, sigma = 0.01)  # Adjust sigma (bandwidth) as needed

# Plot the results
plot(kden, main = "Kernel Density Estimation of Restroom Locations")
```

## Hot Spot Analysis (Getis-Ord Gi):
Hot Spot Analysis using the Getis-Ord Gi* statistic is a spatial statistical method used to identify statistically significant clusters (hot spots or cold spots) of high or low values in geographic data. This technique helps determine areas where a phenomenon (e.g., restroom accessibility) is unusually concentrated or sparse.











