---
title: "visualization_plot"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---
```{r default_setting, include=FALSE}
library(tidyverse)
library(sf)
library(ggplot2)
library(purrr)
library(tidyverse)
library(dplyr)
library(janitor)
library(here)
library(leaflet)
library(reader)
library(plotly)
library(knitr)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r clean_restroom_data,echo = FALSE, message = FALSE, warning = FALSE}
# Import data
restroom_cleaned = read_csv(here::here("./Data/Public_Restrooms_20241203.csv")) %>% 
  janitor::clean_names()
subway_df = read_csv(here::here("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")) %>% 
  janitor::clean_names()

# Clean restroom data
restroom_cleaned = restroom_cleaned %>% 
  select(
    -website, -operator, -hours_of_operation 
  ) %>% 
  rename(
    restroom_latitude = latitude,
    restroom_longitude = longitude,
    restroom_location = location
  ) %>% 
  mutate(
    restroom_latitude = as.numeric(restroom_latitude),
    restroom_longitude = as.numeric(restroom_longitude),
    restroom_location = st_as_sfc(restroom_location), #convert to point
    open = factor(
      open,
      levels = c("Future", "Seasonal", "Year Round"),
      ordered = TRUE
    ),
    accessibility = factor(
      accessibility,
      levels = c("Not Accessible", "Partially Accessible", "Fully Accessible"),
      ordered = TRUE
    ),
    changing_stations = case_when(
      changing_stations %in% c("Yes, in single-stall all gender restroom only",
                                "Yes, in women's restroom only",
                                "Yes") ~ 1,
      changing_stations == "No" ~ 0
    ),
    status = case_when(
      status %in% c("Not Operational",
                    "Closed for Construction",
                    "Closed") ~ 0,
      status == "Operational" ~ 1
    )
  ) 

# Convert dataframe to sf for spatial operations
restroom_sf = st_sf(restroom_cleaned, crs = 4326)

restroom_df = restroom_cleaned %>% 
  filter(location_type == 'Transit')
```


```{r clean_subway_data, echo = FALSE, message = FALSE, warning = FALSE}
# Load the dataset
subway_df = read_csv(here::here("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv"))

# Clean the dataset
subway_cleaned = subway_df %>% 
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    route1, route2, route3, route4, route5, route6, 
    route7, route8, route9, route10, route11, 
    entry, exit_only, vending, entrance_type, ada
  ) %>%  
  mutate(across(starts_with("route"), ~ replace_na(as.character(.), ""))) %>% 
   mutate(
    station_latitude = as.numeric(station_latitude),
    station_longitude = as.numeric(station_longitude)
  )
```

### NYC Public Restrooms 

#### Type of Restroom Location
```{r restroom location_type barplot}
restroom_location_bar = restroom_cleaned |> 
  group_by(location_type) |> 
  count(location_type) |> 
  mutate(loation_type = fct_reorder(location_type, n)) |> 
  plot_ly (x = ~location_type, y = ~n, color = ~location_type,
           type = "bar",text = ~n, textposition = "outside") |> 
  layout(
    xaxis = list(title = "Location Type"),   
    yaxis = list(title = "Number of Restrooms") 
  )

 restroom_location_bar
```
The majority of collected restrooms are located in **_Park_** and **_Library_**, and only a few restrooms are located in **_Privately Owned Public Space_**, **_Public Plaze_**, and **_Transit_**. This might provide guidance for the future establishment of public restrooms: public plaza have siginificant foot traffic, yet public restrooms are so scarce.  

#### Whether or not the Restroom is Only Operational on a Seasonal Schedule
```{r restroom open}
restroom_open_bar = restroom_cleaned |> 
  filter(!is.na(open))|>
  group_by(open) |> 
  count(open, name = "n") |> 
  ungroup()|>
  mutate(open = fct_reorder(open, n)) |> 
  plot_ly (x = ~open, y = ~n, color = ~open,
           type = "bar", text = ~n, textposition = "outside",
           texttemplate = "%{text}") |> 
  layout(
    xaxis = list(title = "Open Type"),   
    yaxis = list(title = "Number of Restrooms") 
  )

 restroom_open_bar

```
From the collected data, New York City provides public restroom facilities as follows, 864 open **_Year Round_**, 123 open **_Seasonal_**, and 12 open **_Future_**. Most public restrooms are open year-round, which is good news!

#### Are Location Type and Open time related?
```{r}
location_type_open <- 
  restroom_cleaned |> 
  filter(!is.na(open)) |> 
  filter(!is.na(location_type))|>
  group_by(location_type, open) |> 
  summarise(count = n(),.groups="drop") |> 
  group_by(open)|>
  mutate(percent = round(count / sum(count) * 100, 1))
 

location_type_open |>
  mutate(hover_text = str_c("Open: ", open, 
                            "<br>Location Type: ", location_type, 
                            "<br>Percent: ", percent, "%")) |> 
  plot_ly(
    x = ~open, 
    y = ~percent, 
    color = ~location_type,
    type = "bar", 
    colors = "viridis", 
    text = ~hover_text, 
    hoverinfo = "text"
  )|>
  layout(
    xaxis = list(title = "Open"),   
    yaxis = list(title = "Percent of Restroom (%)"))
```
From the plot, we can get that the open type is related to location type. All future open-status public restrooms are located in library. All seasonal open-status public restrooms are located in park. This might be due to the impact of climate on the park and its visitor numbers. 

#### The Level of ADA Accessibility where the Restroom is Located
```{r restroom accessibility}
restroom_accessibility_bar = restroom_cleaned |> 
  filter(!is.na(accessibility))|>
  group_by(accessibility) |> 
  count(accessibility, name = "n") |> 
  ungroup()|>
  mutate(accessibility = fct_reorder(accessibility, n)) |> 
  plot_ly (x = ~accessibility, y = ~n, color = ~accessibility,
           type = "bar",text = ~n, textposition = "outside") |> 
  layout(
    xaxis = list(title = "Accessibility Type"),   
    yaxis = list(title = "Number of Restrooms") 
  )

 restroom_accessibility_bar
```
Although the number of **_Fully Accessible_** is the highest, there are still quite a few that are **_Not Accessible_**. We still need to work on improving the implementation of public restrooms to provide services for everyone. 

#### The Number and Gender Designation of the Stalls
```{r restroom type}
restroom_type_bar = restroom_cleaned |> 
  filter(!is.na(restroom_type))|>
  group_by(restroom_type) |> 
  count(restroom_type, name = "n") |> 
  ungroup()|>
  mutate(restroom_type = fct_reorder(restroom_type, n)) |> 
  plot_ly (x = ~restroom_type, y = ~n, color = ~restroom_type,
           type = "bar", text = ~n, textposition = "outside") |> 
  layout(
    xaxis = list(title = "Restroom Type"),   
    yaxis = list(title = "Number of Restrooms")
  )

 restroom_type_bar

```
**_Multi-Stall W/M Restrooms_** are the most commonly built, while **_Multi-Stall All Gender Restrooms_** are the least, with only 4. We need to provide more Multi-Stall All Gender Restrooms so they can serve all groups and more people at same time. 


### NYC Subway

#### Vending machine status
```{r subway vending}
subway_vending_pie = subway_cleaned |> 
  group_by(vending) |> 
  count(vending, name = "n") |> 
  mutate(vending = fct_reorder(vending, n)) |> 
  plot_ly (labels = ~vending, values = ~n, type = "pie",
           textinfo = "label+percent", 
           marker = list(colors = viridis::viridis(2)) )|> 
  layout(title = "Subway Vending"
     )

 subway_vending_pie

```
The pie chart tells that mostly subway stations have vending machines now. 


#### Entrance Type of Subway Station
```{r subway entrance_type}
subway_entrance_bar = subway_cleaned |> 
  group_by(entrance_type) |> 
  count(entrance_type, name = "n") |> 
  mutate(entrance_type = fct_reorder(entrance_type, n)) |> 
  plot_ly (x = ~entrance_type, y = ~n, color = ~entrance_type, colors = "viridis", text = ~n, textposition = "outside", type = "bar") |> 
  layout(
    xaxis = list(title = "Subway Entrance Type"),   
    yaxis = list(title = "Number of Subway Stations") 
  )

 subway_entrance_bar

```
Majority of subway stations' entrance type is **_Stair_**. The least common subway station entrance is **_Walkway_** and **_Ramp_**. 


#### Are Entrance Type and ADA level for Subway Station related?
```{r }
entrance_type_ada <- 
  subway_cleaned |> 
  group_by(entrance_type, ada) |> 
  summarise(count = n(),.groups="drop") |> 
  group_by(entrance_type)|>
  mutate(percent = round(count / sum(count) * 100, 1))

entrance_type_ada |>
  mutate(hover_text = str_c("Entrance Type: ", entrance_type, 
                            "<br>Ada: ", ada, 
                            "<br>Percent: ", percent, "%")) |> 
  plot_ly(
    x = ~entrance_type, 
    y = ~percent, 
    color = ~ada,
    type = "bar", 
    colors = "viridis", 
    text = ~hover_text, 
    hoverinfo = "text"
  )|>
  layout(
    xaxis = list(title = "Entrance Type"),   
    yaxis = list(title = "Percent of Subway Stations (%)"),
    title = "")
```
This plot shows the relationship between entrance type of subway stations and their ADA level. The stations with **_walkway_** or **_elevator_** entrances are more likely to be considered true in ADA level, while stations with **_door_**, **_ramp_**, or **_stair_** entrances are more likely to be considered false in ADA level. However, according to previous plot, we really need to build more **_Walkway_**, **_Easement_**, **_Elevator_** and **_Escalator_** entrances for the subway stations.  
