---
title: "export_clean_data"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(tidyr)
library(sf)
```

# Import Data

We import the restroom data from NYC Open Data: https://data.cityofnewyork.us/City-Government/Public-Restrooms/i7jb-7jku/about_data and subway data from the P8105 course website Homework 2 (Problem 1): https://p8105.com/homework_2.html

Then we use `skimr::skim()` to check variables of both datasets.

```{r import_and_view, message=FALSE}
subway_dirty = read_csv("NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% 
  janitor::clean_names()

restroom_dirty = read_csv("Public_Restrooms_20241203.csv") %>% 
  janitor::clean_names()

# View subway_dirty
skimr::skim(subway_dirty)

# View restroom_dirty
skimr::skim(restroom_dirty)
```
# Clean Variable

```{r clean_subway_data}
subway_cleaned <- subway_dirty %>% 
  mutate(
    entrance_type = factor(
      entrance_type,
      levels = c("Stair", "Easement", "Door", "Walkway", "Escalator", "Ramp", "Elevator")
    )
  ) %>% 
  mutate(
    staffing = factor(
      staffing,
      levels = c("NONE", "Spc Ev", "PART", "FULL")
    )
  ) %>% 
  dplyr::select(
    division, line, station_name, station_latitude, station_longitude, 
    entrance_type, vending, staffing, ada, free_crossover, station_location
  ) %>% 
  mutate(
    x = station_longitude,
    y = station_latitude
  ) %>% 
  st_as_sf(coords = c("x", "y"), crs = 4326) %>% 
  distinct(station_name, .keep_all = TRUE)

# Convert subway data to sf object
subway_sf = st_sf(subway_cleaned, crs = 4326)
```

For subway data, we factor character variables `entrance_type` and `staffing` into categorical variables and select variable like `station_latitude`, `station_longitude`, `ada`, etc for future analysis.

After cleaning, `subway_cleaned` contains `r nrow(subway_cleaned)` rows and `r ncol(subway_cleaned)` columns.

```{r clean_restroom_data}
restroom_cleaned <- restroom_dirty %>% 
  dplyr::select(
    -website, -operator, -hours_of_operation, -additional_notes
  ) %>% 
  rename(
    restroom_latitude = latitude,
    restroom_longitude = longitude,
    restroom_location = location
  ) %>% 
  mutate(
    restroom_latitude = as.numeric(restroom_latitude),
    restroom_longitude = as.numeric(restroom_longitude),
    restroom_location = st_as_sfc(restroom_location),
    restroom_open = factor(
      open,
      levels = c("Future", "Seasonal", "Year Round")
    ),
    restroom_accessibility = factor(
      accessibility,
      levels = c("Not Accessible", "Partially Accessible", "Fully Accessible")
    ),
    restroom_changing_stations = case_when(
      changing_stations %in% c("Yes, in single-stall all gender restroom only",
                                "Yes, in women's restroom only",
                                "Yes") ~ 1,
      changing_stations %in% c("No", NA) ~ 0
),
    restroom_status = case_when(
      status %in% c("Not Operational",
                    "Closed for Construction",
                    "Closed") ~ 0,
      status == "Operational" ~ 1
    )
  ) %>% 
  dplyr::select(
    -open, -accessibility, -changing_stations, -status
  )

# Convert dataframe to sf for spatial operations
restroom_sf <- st_as_sf(restroom_cleaned, crs = 4326)

# Filter restrooms near transit
restroom_near_transit <- restroom_cleaned %>% 
  filter(location_type == 'Transit')

subway_with_restrooms <- st_join(subway_sf, restroom_sf, join = st_nearest_feature)
```

For restroom data, we factor the character variables (`status`, `open`, `accessibility`, `changing_stations`) into binary and categorical variables.

After cleaning, `restroom_cleaned` contains `r nrow(restroom_cleaned)` rows and `r ncol(restroom_cleaned)` columns.

Then we merge the file by enhancing both dataframes to `sf` and find the closest `restroom` subject to each `subway` subjects.

# Export clean data

```{r}
save_restroom_path = "cleaned_restroom_data.csv"
write_csv(restroom_sf, save_restroom_path, na="")

save_subway_path = "cleaned_subway_data.csv"
write_csv(subway_sf, save_subway_path, na="")

save_subway_restroom_path = "cleaned_subway_restroom_data.csv"
write_csv(subway_with_restrooms, save_subway_restroom_path, na="")